---
title: "Using gs.engine(): Genstat Socket Engine"
author: "Yuxiao Wang"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
  toc: true
  fig_caption: true
vignette: >
  %\VignetteIndexEntry{Using gs.engine(): Genstat Socket Engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo = FALSE}
img {
    max-width: 50%;  
    height: auto; 
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(gsengine)
knitr::knit_engines$set(gs = gsengine::gs.engine())
```



# Overview
This vignette demonstrates how to use the gs.engine function provided by the gsengine package to run Genstat code chunks from within an R Markdown document.

It assumes a Genstat server is already running and accessible over a socket connection.



# Setting up the connection
You can configure the Genstat server host and port via environment variables:
```{r}
Sys.setenv(GENSTAT_HOST = "sc389508.UoA.auckland.ac.nz")
Sys.setenv(GENSTAT_PORT = "8085")
knitr::knit_engines$set(gs = gsengine::gs.engine())
```



# Writing Genstat chunks
To **run Genstat code** in your R Markdown file, specify the chunk engine as 'gs':

    {gs}

Example:
```{gs}
VARIATE [VALUES=1,2,3] Foo
```

```{gs}
PRINT Foo
```
This code will be sent to the Genstat server, and any output will be captured and rendered in the final document.



# Echo-only Code Blocks

1) If you want to display the Genstat code **without evaluating** it(eval=FALSE):
    
    {gs, eval=FALSE}
  
    Example:
    
    ```{gs, eval=FALSE}
    PRINT 'This should not run.'
    ```


2) By default, the 'gs.engine' **does** show the Genstat commands submitted in each chunk.

    To **hide Genstat code lines** in the rendered output, use the 'echo = FALSE' option in the chunk header: 
    
    {gs, echo=FALSE}
  
    Example:
    ```{gs, echo=FALSE}
    PRINT 'This should run.'
    ```

    This will only include the Genstat result or tables in the final document output.

    This setting behaves similarly to R Markdown’s native echo = TRUE option for R code chunks.



# Displaying Genstat Graphics
If Genstat outputs a **PNG image**, it will be displayed directly in the output document. These images are saved locally with filenames like 'img_1.png', 'img_2.png', etc.

Example:
```{gs}
DGRAPH !(1,2,3,4,5); !(10,15,7,20,12)
```

# Importing data into Genstat
Importing data into Genstat can often have a side-effect in that messages may be returned. Whilst we can clearly just
echo the text, it might be useful to, for example, turn table data into tables. The messages (and output) are controlled 
by the a `PRINT` option, which can be one of `CATALOGUE`, `SUMMARY`, or `*`. If not specified, then the default is `[PRINT=CATALOGUE]`.
*Note*: these options may be shortened to a single character.

Let's look at the output we get from each option

```{gs}
IMPORT '%DATA%/Iris.gsh'
```
```{gs}
IMPORT [PRINT=C] '%DATA%/AbaloneAge.gsh'
```

```{gs}
IMPORT [PRINT=SUMMARY] '%DATA%/Iris.gsh'
```

```{gs}
IMPORT [PRINT=*] '%DATA%/Iris.gsh'
```
As we can see, there are three (or four depending on how you look at it) possibilities. The last option `[PRINT=*]` 
suppresses output. `[PRINT=CATALOG]` gives a table summary which gives the order of the variables, the variable type,
the number of observations for each variable, and the variable name. `[PRINT=SUMMARY]` gives a simple numerical summary (minimum, mean, maximum, $n$, $n_{missing}$)
of each variable of type `variate`, and of each `factor` ($n$, $n_{missing}$, levels).

```{gs,store_complex_tables=TRUE, store_simple_tables=TRUE}
IMPORT [PRINT=SUMMARY] '%DATA%/Iris.gsh'
```

We might regard 



# Warning and Table Formatting
The 'gs.engine' will automatically:

1) **Highlight Genstat warnings/Faults** in red.

2) **Format complex tables** (like factors/variates) as Markdown tables.

3) **Format simple tables** (like value/rank summaries) as well-structured Markdown tables.

4) **Recognize and format structural markers** in Genstat output for better readability:
  
    -Section breaks such as 1., 2., or 3.1 are rendered as Markdown section dividers:---

    -Headings indicated by === (equal signs under a line) will be converted into bold headings.

    -Subheadings indicated by --- (hyphens under a line) will be converted into italicized subheadings.

    Example:
```{gs, store_complex_tables=TRUE, store_simple_tables=TRUE}
IMPORT '%DATA%/WheatTrials.xlsx'; SHEET='Trial A1'
TREATMENTS Cultivar
BLOCKS Rep
ANOVA [FPROB=yes; PSE=LSD] Mean_Population_m2
AMCOMPARISON Cultivar
ANOVA [FPROB=yes; PSE=LSD] Yield_14%_t_ha_adj
AMCOMPARISON Cultivar
```


You can enable **table storage** by setting options in the chunk header:

    {gs, store_complex_tables=TRUE, store_simple_tables=TRUE}
  
Tables will be available as lists in global variables:
  
    complex_tables_all_chunks
  
    simple_tables_all_chunks
  
Example:
```{r}
complex_tables_all_chunks
simple_tables_all_chunks
```



# Troubleshooting
If you experience connection issues:

1. Ensure your Genstat socket server is running and listening on the correct port.

2. Test using 'telnet <host> <port>' or similar tools.



# Session Info
```{r}
sessionInfo()
```

